---
title: "PA1_template"
author: "Valter Lacerda"
date: "Saturday, August 16, 2014"
output: html_document
---
# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data

Reading the data after unzipped the data file "activity.zip".

```{r}
# read in the CSV file
activity <- read.csv("activity.csv")
```

Now, the answers to the questions.

## What is mean total number of steps taken per day?

-> Ignore the missing values in the dataset

-> Histogram of the total number of steps taken each day
-> Calculation of the mean and median total number taken each day.

```{r histogram1}
# create a data frame containing the sum of steps for each date
totalstepsperday <- data.frame(tapply(activity$steps, activity$date, sum, na.rm=TRUE))

names(totalstepsperday) <- "steps"

library(ggplot2)
ggplot(totalstepsperday) + aes(x=steps) + geom_histogram(binwidth=1000) + labs(title="Histogram of Total Number of Steps Per Day")

# calculate the mean and median
mean(totalstepsperday$steps)
median(totalstepsperday$steps)
```

-> We can notice tha the total number of steps per day is 9354 (mean), and the median total number of steps per day is 10395.

## What is the average daily activity pattern?

-> Time series plot of the average number of steps taken during each 5-minute interval

```{r timeseries1}
# create a data frame
avgstepsperinterval <- data.frame(tapply(activity$steps, activity$interval, mean, na.rm=TRUE))

names(avgstepsperinterval) <- "steps"

avgstepsperinterval$interval <- as.integer(row.names(avgstepsperinterval))

ggplot(avgstepsperinterval) + aes(x=interval, y=steps) + geom_line() + labs(title="Time Series Plot of Average Number of Steps Per Interval")

# determine which interval contains the maximum number of steps
names(which.max(avgstepsperinterval$steps))
```

-> The interval with the maximum number of steps is 835.

## Imputing missing values

```{r}
# calculate the number of missing values
sum(is.na(activity$steps))
```

-> There are 2304 missing values for "steps" in the dataset.

-> Here I create a new dataset (activity.missing) where I have imputed values (the mean for that 5-minute interval) for the missing values of "steps"

```{r}
# create the new dataset
activity.missing <- activity

# loop through each row in activity
for (i in 1:nrow(activity)) {
    # check if the "steps" value is missing
    if(is.na(activity$steps[i])) {
        # if TRUE, replace the missing value with the mean for that interval
        activity.missing$steps[i] <- avgstepsperinterval$steps[which(avgstepsperinterval$interval==activity$interval[i])]
    }
}
```

-> Histogram of the total number of steps taken each day, along with a calculation of the mean and median total number taken each day.

```{r histogram2}
# create a data frame containing the sum of steps for each date
totalstepsperday.imp <- data.frame(tapply(activity.missing$steps, activity.missing$date, sum, na.rm=TRUE))

names(totalstepsperday.imp) <- "steps"

ggplot(totalstepsperday.imp) + aes(x=steps) + geom_histogram(binwidth=1000) + labs(title="Histogram of Total Number of Steps Per Day")

# calculate the mean and median total steps per day
mean(totalstepsperday.imp$steps)
median(totalstepsperday.imp$steps)
```

-> For the new dataset, the mean and median total number of steps per day are both 10766. This is different from the first part of the assignment. 

## Are there differences in activity patterns between weekdays and weekends?

-> To answer this question, let's check for differences in activity patterns between weekdays and weekends.

```{r}
# The wday function from lubridate determines if each date is weekday or weekend
library(lubridate)
activity.missing$daytype <- as.factor(ifelse(wday(activity.missing$date, label=TRUE) %in% c("Sat", "Sun"), "weekend", "weekday"))
```

-> Spliting the dataset by daytype.

```{r}
# split activity.missing by "daytype"
s <- split(activity.missing, activity.missing$daytype)

# create data frames containing the mean number of steps for each interval
avgsteps.weekday <- data.frame(tapply(s$weekday$steps, s$weekday$interval, mean))
avgsteps.weekend <- data.frame(tapply(s$weekend$steps, s$weekend$interval, mean))

# fix the variable names
names(avgsteps.weekday) <- "steps"
names(avgsteps.weekend) <- "steps"

# add new variables for the interval
avgsteps.weekday$interval <- as.integer(row.names(avgsteps.weekday))
avgsteps.weekend$interval <- as.integer(row.names(avgsteps.weekend))

# add new variables for the daytype
avgsteps.weekday$daytype <- factor(rep("weekday", nrow(avgsteps.weekday)), levels=c("weekday", "weekend"))
avgsteps.weekend$daytype <- factor(rep("weekend", nrow(avgsteps.weekend)), levels=c("weekday", "weekend"))
```

-> Comparing the weekday and weekend datasets.

```{r timeseries2,fig.width=10}
# combine the weekday and weekend datasets
avgsteps.combined <- rbind(avgsteps.weekday, avgsteps.weekend)

ggplot(avgsteps.combined) + aes(x=interval, y=steps) + facet_grid(daytype ~ .) + geom_line() + labs(title="Time Series Plot of Average Number of Steps Per Interval: Comparing Weekday (top) vs. Weekend (bottom)", y="Number of steps")
```

-> As we can notice that there are differences in activity patterns between weekdays and weekends.
